Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Model

### 2025-10-23 analysis/paper/access-ice.qmd
Adds descriptions to zenodo list

```{r datasets}
datasets <- list(cdr = CDR(), osi = OSI())
which_dataset <- "cdr"
```

@@

```{r hindcast_extent}
hindcast_extent <- here::here("data/derived/hindcast_extent.csv") |>
  zenodo() |>
  fread() |>
  _[, model := relevel(factor(model), "S2")] |>
  _[, lag := as.numeric(as.Date(time) - as.Date(forecast_time))] |>
  _[lag > 0] |>
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")] # This member is wrong
```

@@

aice_forecast_example <- c(
  lapply(c("S1", "S2"), \(model) {
    month_clim <- formatC(month(forecast_date), width = 2, flag = "0")
    clim <- here::here(glue::glue(
      "data/derived/climatology/{model}/{month_clim}/em.nc"
    )) |>
      zenodo()
    hindcast(forecast_date, model, 1) |>
      cdo_ydaysub(clim) |>
      cdo_execute() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := model]
  }),
  list(
    CDR() |>
      anomalies() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := "cdr"]
  )
) |>
  rbindlist()

thickness_example <- c(
  S2 = "/g/data/ux62/access-s2/hindcast/raw_model/ice/hi/daily/e01/di_hi_{format(forecast_date, '%Y%m%d')}_e01.nc",
  S1 = "/g/data/ub7/access-s1/hc/raw_model/unchecked/ice/ice/daily/e01/di_ice_{format(forecast_date, '%Y%m%d')}_e01.nc"
) |>
  zenodo() |>
  purrr::imap(\(f, model) {
    f |>
      glue::glue() |>
      cdo_selname("hi") |>
      remap_cdr() |>
      cdo_execute(cache = TRUE) |>
      ReadNetCDF("hi", subset = list(time = sample_date))
  }) |>
  rbindlist(idcol = "model")

@@

```{r mean_thickness_funs}

# mean_thickness <- function(concentration, thickness, cell_area = NULL) {
#   rcdo::cdo_cache_set(here::here("data/temp/cache/"))
#   if (is.null(cell_area)) {
#     cell_area <- thickness |>
#       rcdo::cdo_gridarea()
#   }

#   volume <- thickness |>
#     rcdo::cdo_mul(concentration) |>
#     rcdo::cdo_mul(cell_area) |>
#     rcdo::cdo_fldsum()

#   area <- concentration |>
#     rcdo::cdo_mul(cell_area) |>
#     rcdo::cdo_fldsum()

#   rcdo::cdo_div(volume, area) |>
#     rcdo::cdo_execute(options = "-L")
# }

# mean_thickness_access <- \(forecast_time, model) {
#   concentration <- hindcast(
#     forecast_time,
#     model = model,
#     members = "em",
#     variable = "aice"
#   )
#   thickness <- hindcast(
#     forecast_time,
#     model = model,
#     members = "em",
#     variable = "hi"
#   )

#   if (length(thickness) == 0) {
#     return(NA_character_)
#   }
#   if (length(concentration) == 0) {
#     return(NA_character_)
#   }

#   cell_area <- thickness |>
#     rcdo::cdo_gridarea()

#   mean_thickness(concentration, thickness, cell_area)
# }
```

```{r hindcast_thickness}
# hindcast_thickness <- CJ(model = c("S2", "S1")) |>
#   _[, .(forecast_time = get_forecast_times(model)), by = model] |>
#   _[mday(forecast_time) == 1] |>
#   _[,
#     concentration := hindcast(
#       forecast_time,
#       model,
#       members = "em",
#       variable = "aice"
#     ) |>
#       zenodo(),
#     by = .(model)
#   ] |>
#   _[mday(forecast_time) == 1] |>
#   _[,
#     thickness := hindcast(
#       forecast_time,
#       model,
#       members = "em",
#       variable = "hi"
#     ) |>
#       zenodo(),
#     by = .(model)
#   ] |>
#   na.omit() |>
#   _[,
#     thickness := unlist(mirai_map(
#       data.frame(concentration, thickness),
#       mean_thickness
#     )[.progress])
#   ] |>
#   na.omit()

@@

```{r errors}
errors <- here::here("data/derived/rmse.Rds") |>
  zenodo() |>
  readRDS() |>
  _[value < 0.01, value := NA] |>
  _[value > 3e13, value := NA] |>
  _[, time := as.Date(time)] |>
  _[, lag := as.numeric(time - time_forecast)] |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[lag > 0] |>
  _[!(month(time_forecast) == 1 & member == "07" & version == "S1")] # This member is wrong
```

@@

```{r topo_lonlat}
topo_lonlat <- here::here("data/raw/ETOPO.nc") |>
  zenodo() |>
  ReadNetCDF() |>
  na.omit()
```

```{r}
clim_std_lon <- datasets |>
  lapply(\(x) {
    x |>
      anomalies(climatology(x)) |>
      cdo_sqr() |>
      cdo_remapmean("r15x360") |>
      cdo_mermean() |>
      cdo_sqrt() |>
      cdo_execute() |>
      ReadNetCDF(c(climatology = "aice")) |>
      _[, let(lat = NULL, lon = lon + 12)]
  }) |>
  rbindlist(idcol = "obs_dataset") |>
  _[, time := as.Date(time)] |>
  _[climatology == 0, climatology := NA] |>
  _[climatology < 0.05, climatology := NA]

@@

```{r rmse_lon_mean}
rmse_lon_mean <- here::here("data/derived/rmse_lon.Rds") |>
  zenodo() |>
  readRDS() |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[, lon := lon + 12] |>
  _[, time := update(time, hour = 0)] |>
  _[, time := as.Date(time)]
```

>>>>

```{r datasets}
datasets <- list(
  cdr = CDR() |>
    zenodo("NSIDC CDR sea-ice concentration."),
  osi = OSI() |>
    zenodo("OSI sea-ice concentration.")
)
which_dataset <- "cdr"
```

@@

```{r hindcast_extent}
hindcast_extent <- here::here("data/derived/hindcast_extent.csv") |>
  zenodo("Sea-ice extent from hindcast.") |>
  fread() |>
  _[, model := relevel(factor(model), "S2")] |>
  _[, lag := as.numeric(as.Date(time) - as.Date(forecast_time))] |>
  _[lag > 0] |>
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")] # This member is wrong
```

@@

aice_forecast_example <- c(
  lapply(c("S1", "S2"), \(model) {
    month_clim <- formatC(month(forecast_date), width = 2, flag = "0")
    clim <- here::here(glue::glue(
      "data/derived/climatology/{model}/{month_clim}/em.nc"
    )) |>
      zenodo("{model} sea-ice concentration climatology for {month_clim}")

    hindcast(forecast_date, model, 1) |>
      zenodo("{model} hindcast for initialised on {forecast_date}") |>
      cdo_ydaysub(clim) |>
      cdo_execute() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := model]
  }),
  list(
    CDR() |>
      anomalies() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := "cdr"]
  )
) |>
  rbindlist()

thickness_example <- c(
  S2 = hindcast(forecast_date, "S2", members = 1, variable = "hi") |>
    zenodo(
      "S2 sea-ice thickness hindcast for {forecast_date} (1st ensemble member)"
    ),
  S1 = hindcast(forecast_date, "S1", members = 1, variable = "hi") |>
    zenodo(
      "S1 sea-ice thickness hindcast for {forecast_date} (1st ensemble member)"
    )
) |>
  purrr::imap(\(f, model) {
    f |>
      glue::glue() |>
      cdo_selname("hi") |>
      remap_cdr() |>
      cdo_execute(cache = TRUE) |>
      ReadNetCDF("hi", subset = list(time = sample_date))
  }) |>
  rbindlist(idcol = "model")

@@

@@

```{r errors}
errors <- here::here("data/derived/rmse.Rds") |>
  zenodo("Hindcast RMSE of sea-ice concentration anomalies") |>
  readRDS() |>
  _[value < 0.01, value := NA] |>
  _[value > 3e13, value := NA] |>
  _[, time := as.Date(time)] |>
  _[, lag := as.numeric(time - time_forecast)] |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[lag > 0] |>
  _[!(month(time_forecast) == 1 & member == "07" & version == "S1")] # This member is wrong
```

@@

```{r topo_lonlat}
topo_lonlat <- here::here("data/raw/ETOPO.nc") |>
  zenodo("ETOPO topography") |>
  ReadNetCDF() |>
  na.omit()
```

```{r clim_std_lon} 
clim_std_lon <- datasets |>
  lapply(\(x) {
    x |>
      anomalies(climatology(x)) |>
      cdo_sqr() |>
      cdo_remapmean("r15x360") |>
      cdo_mermean() |>
      cdo_sqrt() |>
      cdo_execute() |>
      ReadNetCDF(c(climatology = "aice")) |>
      _[, let(lat = NULL, lon = lon + 12)]
  }) |>
  rbindlist(idcol = "obs_dataset") |>
  _[, time := as.Date(time)] |>
  _[climatology == 0, climatology := NA] |>
  _[climatology < 0.05, climatology := NA]

@@

```{r rmse_lon_mean}
rmse_lon_mean <- here::here("data/derived/rmse_lon.Rds") |>
  zenodo("Hindcast RMSE in latitude slices") |>
  readRDS() |>
  _[version == "persistence", member := "0em"] |>
  _[member == "0em"] |>
  _[, lon := lon + 12] |>
  _[, time := update(time, hour = 0)] |>
  _[, time := as.Date(time)]
```
